---
title: "Untitled"
format: html
---

### Analysis of the blood data from the AGILE trial
### analysis is blinded, with treatments named "guinness" and "corona"
```{r}
library(CATALYST)
library(harmony)
library(cowplot)
library(flowCore)
library(scater)
library(SingleCellExperiment)
library(viridis)
library(RColorBrewer)
library(ggforce)
library(ggplot2)
library(ggrepel)
library(ggsci)
library(data.table)
library(emmeans)
library(dplyr)
library(tidyr)
library(ggsignif)
library(rstatix)
library(multcomp)
library(scales)
library(ggthemes)
library(slingshot)
library(ggbeeswarm)
library(ggbreak)
library(DescTools)
library(pak)
library(gghighlight)
library(scattermore)
library(clustree)
library(gridExtra)
library(patchwork)
library(ggbump)
library(stringr)
library(grid)
```

```{r}
"c:/Users/dan_m/OneDrive - University College London/UCL_Senior_Research_Fellow/AGILE/Blood"
```

## loads in FCS files and create flowset object
### list fcs files in folder, choose ones requied and then run read.flowset
```{r}
fcs <- list.files(pattern = ".fcs$")
fcs
fs <- read.flowSet(fcs, transformation = F, truncate_max_range = F)
```


## setting up flourophore information
### check columns (fluorophores) and remove any not needed
### when running for the first time, save the columns into an excel to form the basis of the fluorophore metadata
### if this has been created before, just load it in
```{r}
all_cols <- colnames(fs)
all_cols
fs <- fs[,-c(1,2,31,35)]
colnames(fs)
cols <- colnames(fs)
cols <- as.data.frame(cols)
#write.csv(cols, file = "Marker.metadata.csv")
markers <- read.csv("Marker.metadata.csv")
markers
```

## setting up sample metadata
### use the file names as the basis for the sample metadata
### first add a column to make the sample id (removes excess characters at end of each fcs file name)
#### more specifically, this drops the last 18 characters (which are the redundant parts of the file name)
### use separate to add individual metadata features - due to "stained_blood" being in the naame, i tell this to become "drop" (because i'll remove it) and "tissue"
## if script run previously then just load it in
```{r}
names <- data.frame(fcs)
sample_df <- names %>% mutate(sample_id = str_sub(fcs, 1, nchar(fcs)-18))
sample_df <- separate(sample_df, col = sample_id, into = c("treatment", "donor", "drop", "tissue", "time"), remove = F, sep = "_")
sample_df <- sample_df[, c(-5)]
#write.csv(sample_df, file = "sample.metadata.csv")
samples <- read.csv("sample.metadata.csv")
samples <- samples[-1]
```

## making the singlecellexperiment object
### double check flourophore and sample_ids match their respective metadata files
### use prepdata to make the sce - here cofactors can be set, which are assessed afterwards
### this step is repeated until final cofactors are decided
```{r}
setdiff(colnames(fs), markers$fluorophore)
setdiff(fcs, samples$fcs)

sce <- prepData(x = fs, panel = markers, md = samples, FACS = T, transform = T, 
  cofactor = c("AF647-A"=1000,"APC-Cy7-A"=1000,"APC-Fire810-A"=1000,"BUV395-A"=1000,"BUV496-A"=1000,     "BUV563-A"=1000,"BUV737-A"=1000,"BUV805-A"=1000,"BV421-A"=1000,"BV480-A"=1000,"BV605-A"=1000,"BV711-A"=1000,"BV750-A"=1000,"BV785-A"=1000,"FITC-A"=1000,"PE-A"=1000,"PE-Cy5-A"=1000,"PE-Cy7-A"=1000,"PE-Fire810-A"=1000,"R718-A"=1000,"RB545-A"=1000,"RB744-A"=1000,"RB780-A"=1000,"SB645-A"=1000,"SBB580-A"=1000,"SBUV605-A"=1000,"SBV570-A"=1000,"V450-A"=1000,"[RB613]-A"=1000,"[RB705]-A"=1000,"[RY610]-A"=1000), 
  panel_cols = list(channel = "fluorophore", antigen = "antigen"),
  md_cols = list(file = "fcs", id = "sample_id",
  factors = c("donor", "treatment", "tissue", "time")))
```

## downsampling
### there is ~13 million cells, which is far more than necessary and is computationally limiting
### therefore downsampling to 5% of cells, applied to each sample individually
### this function is straight from chatgpt, but seems to work well
```{r}
downsample <- function(sce, sample_col = "sample_id", proportion = 0.05, min_cells = 10) {
  # Get sample labels
  samples <- colData(sce)[[sample_col]]
  
  # For each sample, randomly select proportion of cells
  selected_cells <- unlist(lapply(unique(samples), function(s) {
    cells_in_sample <- which(samples == s)
    n <- max(round(length(cells_in_sample) * proportion), min_cells)  # enforce a minimum
    n <- min(length(cells_in_sample), n)  # avoid oversampling
    sample(cells_in_sample, n)
  }))
  
  # Subset the SCE to those cells
  sce_downsampled <- sce[, selected_cells]
  return(sce_downsampled)
}
numbers_before <- table(sce$sample_id)
sce <- downsample(sce, sample_col = "sample_id", proportion = 0.05)
numbers_after <- table(sce$sample_id)
```
## assess cofactors
### check how many cells in experiment to get an idea of computational load
### assess cofactors - looking for negative peak to be around 0 whilst maintaining a separation from positive peak
### there is a level of subjectivity
```{r}
table(sce$treatment)
plotExprs(sce, color_by = "donor")
```

## Batch correction
### use the harmony package to apply batch correction - using donor as the correction
```{r}
expdata <- sce@assays@data$exprs
meta <- sce@colData$donor
expdata <- t(expdata)
nrow(expdata)
nrow(meta)
harmony_output <- HarmonyMatrix(expdata, meta_data = meta, do_pca = F)
sce@assays@data$uncorrected <- sce@assays@data$exprs
sce@assays@data$exprs <- t(harmony_output)
```
## clustering
## run the clustering and assess elbow plot - this will determine appropriate k number for selecting resolution
```{r}
set.seed(61)
sce <- CATALYST::cluster(sce, features = NULL, xdim = 10, ydim = 10, maxK = 30, verbose = T, seed = 61)

delta <- delta_area(sce)
delta <- delta$data
ggplot(delta, aes(x = k, y = y))+geom_point(size = 5)+geom_line(linewidth = 1.5)+
  theme_classic()+
  theme(axis.text = element_text(colour = "black", size = 18), 
        axis.title = element_text(size = 18, colour = "black"))+
  xlab("K (number of clusters)")+ylab("Relative change in \n area under CDF curve")

```

### create umap
```{r}
sce <- runDR(sce, "UMAP", features = NULL)
```

## save/load here
```{r}
#saveRDS(sce, file = "agile_blood_clustered.rds", compress = F)
sce <- readRDS("agile_blood_clustered.rds")
```

## inspect umap by metadata - this will confirm absence of batch effects and generally view how groups behave
```{r}
variables <- c("donor", "treatment", "time")
umaps <- list()
for (i in variables) {
umaps[[i]] <- plotDR(sce, "UMAP", color_by = i)+theme_bw()+
    geom_scattermore()
} 
p <- plot_grid(umaps$donor, umaps$treatment, umaps$time)
p
```

## if it looks reasonable (i.e doesn't split by donor, suggesting batche effect) start assessing the resolution (also known as meta or k)
### here i use the previously generated delta plot to get an idea of what an appropriate k might be - i then also "over-cluster" on purpose
## i also plot the expression of all markers in the umap. whilst this is computationally intensive and takes a while, it is worth it as it allows us to see if a marker is expressed across clusters or is distinct to one cluster

```{r}
res <- c("meta7", "meta10", "meta12", "meta15")
umaps <- list()

for (i in res) {
umaps[[i]] <- plotDR(sce, "UMAP", color_by = i)+theme_bw()+
    geom_scattermore()
} 

p1 <- plot_grid(umaps$meta7, umaps$meta10, umaps$meta12, umaps$meta15)
p1

ggsave(plot = p1, filename = "umaps_resolution_comparison.png", dpi = 500,
height = 6, width = 16)

markers.input <- markers$antigen
p2 <- plotDR(sce, color_by = markers.input)+theme_bw()
ggsave(plot = p2, filename = "umap_all_markers.png", dpi = 500,
height = 16, width = 16, bg = "white")
```